Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-12-11 rnni_convexity.tex
added RNNI diameter proof + a few more tweaks (show keys too)

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{float}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage[noend]{algorithmic}
\usepackage{todonotes}
\usepackage{enumitem}
\usepackage[misc]{ifsym}
\usepackage[foot]{amsaddr}
\usepackage[hidelinks]{hyperref}
\usepackage{csquotes}
\usepackage[style=authoryear,ibidtracker=false,uniquename=false,giveninits=true,terseinits=true,maxbibnames=5,backend=biber]{biblatex}
\renewbibmacro{in:}{}
\addbibresource{rnni_convexity.bib}

@@

\newtheorem{definition}{Definition}
\newtheorem{theorem}[definition]{Theorem}
\newtheorem{conjecture}[definition]{Conjecture}
\newtheorem{lemma}[definition]{Lemma}
\newtheorem{corollary}[definition]{Corollary}
\newtheorem{proposition}[definition]{Proposition}

@@

\begin{abstract}
In this paper we study the graph of ranked phylogenetic trees where the adjacency relation is given by a local rearrangement of the tree structure.
Our work is motivated by tree inference algorithms, such as maximum likelihood and Markov Chain Monte Carlo methods, where the geometry of the search space plays a central role for efficiency and practicality of the optimisation.
We hence focus on understanding the geometry of the space (graph) of ranked trees, the so-called ranked nearest neighbour interchange ($\rnni$) graph.
For example, we find exact estimates of the radius and diameter of the space.
Since the $\rnni$ graph is a generalisation of the classical nearest neighbour interchange ($\nni$) graph to ranked phylogenetic trees, we compare geometric and algorithmic properties of the two graphs.
Surprisingly, we discover that both geometric and algorithmic properties of $\rnni$ and $\nni$ are quite different.
For example, we establish convexity of certain natural subspaces in $\rnni$, which are not convex is $\nni$.
Furthermore, our results suggest that the complexity of computing distances in the two graphs is different.
\end{abstract}

@@

\todo{This format takes up too much space.
Why do we need END statements?}
\begin{algorithm}[H]
\caption{$\findpath$($T,R$)}
\label{alg:find_path}
\begin{algorithmic}[1]
\STATE $T' := T$, $p := [T']$
\FOR {$k = 1, \dots, n-2$}
  \STATE Let $C_k$ be the cluster induced by the node with rank $k$ in $R$ \label{alg:find_path:line:cluster}
  \WHILE {$\rank_{T'}(\mrca_{T'}(C_k))>k$}
    \STATE Update $T'$: Decrease $\rank_{T'}(\mrca_{T'}(C_k))$ by an $\rnni$ move \label{alg:findpath:line:move_set_down}
    \STATE $p = p+T'$
  \ENDWHILE
\ENDFOR
\RETURN $p$
\end{algorithmic}
\end{algorithm}

@@

\textcite[Theorem~7]{Gavryushkin2018-ol} gave an upper bound for the diameter of $\rnni$ space.
In this section we improve that result and calculate the exact diameter and radius of the $\rnni$ space.
To achieve this we will employ our results on caterpillar trees from the previous sections.

The results of previous section imply that $d([1,2,3,\ldots,n], [n-1,n,n-2,n-3, \ldots, 1]) = \frac{(n-1)(n-2)}{2}$, which is the maximum possible distance between caterpillar trees.

\begin{corollary}
It is $\Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$.
\label{corollary:diameter}
\end{corollary}

\proof
\todo{Why exactly is this true for arbitrary, not necessarily caterpillar, trees?}
\endproof

We now proceed to show that the radius of the $\rnni$ graph equals its diameter.
Therefore we use the algorithm $\mdtree$ that has been introduced in Section~\ref{section:alg_mdtree}.
Note that the results provided in this section are only true if the input tree of $\mdtree$ is a caterpillar tree.

\begin{lemma}
Let $T$ be a caterpillar tree.
Any tree $R = \mdtree(T)$ has distance $d(T,R) = \Delta(\rnni)$ to $T$.
\label{lemma:max_dist_caterpillar}
\end{lemma}

\begin{proof}
Recall that with Corollary~\ref{corollary:diameter} we know that $\Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
Within this proof we assume, without loss of generality, that it is $T = [1,2,3,\ldots,n]$.
We prove the lemma by induction on the number of taxa $n$.

@@

In Section~\ref{alg:max_dist_tree} we already discussed that $\mdtree$ is a non-deterministic algorithm.
In fact, the output tree of $\mdtree$ can have any tree topology, since in each step of the algorithm the next taxon can be added on any edge incident to a leaf in the tree.
Therefore, if the input tree is a caterpillar tree, $\mdtree$ can compute a tree with any tree topology that has maximum distance from the input tree.
And as permuting labels of trees does not change the distance between them, there exists for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
By this we obtain Corollary~\ref{corollary:radius}.

\begin{corollary}
The radius of the $\rnni$ graph is $\rad(\rnni) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
\label{corollary:radius}
\end{corollary}

>>>>

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{float}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage[noend]{algorithmic}
\usepackage{todonotes}
\usepackage{enumitem}
\usepackage[misc]{ifsym}
\usepackage[foot]{amsaddr}
\usepackage[hidelinks]{hyperref}
\usepackage{csquotes}
\usepackage[notref,notcite]{showkeys}
\usepackage[style=authoryear,ibidtracker=false,uniquename=false,giveninits=true,terseinits=true,maxbibnames=5,backend=biber]{biblatex}
\renewbibmacro{in:}{}
\addbibresource{rnni_convexity.bib}

@@

\newtheorem{definition}{Definition}
\newtheorem{theorem}[definition]{Theorem}
\newtheorem{conjecture}[definition]{Conjecture}
\newtheorem{lemma}[definition]{Lemma}
\newtheorem{proposition}[definition]{Proposition}

@@

\begin{abstract}
In this paper we study the graph of ranked phylogenetic trees where the adjacency relation is given by a local rearrangement of the tree structure.
Our work is motivated by tree inference algorithms, such as maximum likelihood and Markov Chain Monte Carlo methods, where the geometry of the search space plays a central role for efficiency and practicality of the optimisation.
We hence focus on understanding the geometry of the space (graph) of ranked trees, the so-called ranked nearest neighbour interchange ($\rnni$) graph.
For example, we find exact estimates of the radius and diameter of the space, improving the best previously known estimates.
Since the $\rnni$ graph is a generalisation of the classical nearest neighbour interchange ($\nni$) graph to ranked phylogenetic trees, we compare geometric and algorithmic properties of the two graphs.
Surprisingly, we discover that both geometric and algorithmic properties of $\rnni$ and $\nni$ are quite different.
For example, we establish convexity of certain natural subspaces in $\rnni$, which are not convex is $\nni$.
Furthermore, our results suggest that the complexity of computing distances in the two graphs is different.
\end{abstract}

@@

\begin{algorithm}[H]
\caption{$\findpath$($T,R$)}
\label{alg:find_path}
\begin{algorithmic}[1]
\STATE $T' := T$, $p := [T']$
\FOR {$k = 1, \dots, n-2$}
  \STATE Let $C_k$ be the cluster induced by the node with rank $k$ in $R$ \label{alg:find_path:line:cluster}
  \WHILE {$\rank_{T'}(\mrca_{T'}(C_k))>k$}
    \STATE Update $T'$: Decrease $\rank_{T'}(\mrca_{T'}(C_k))$ by an $\rnni$ move \label{alg:findpath:line:move_set_down}
    \STATE $p = p+T'$
  \ENDWHILE
\ENDFOR
\RETURN $p$
\end{algorithmic}
\end{algorithm}

@@

\textcite[Theorem~7]{Gavryushkin2018-ol} gave an upper bound for the diameter of $\rnni$ space, which we recall is denoted by $\Delta(\rnni)$.
They showed that $\Delta(\rnni) = n^2 - 3n - 5/8$.
In this section we improve that result and calculate the exact diameter and radius of the $\rnni$ space.

\begin{theorem}
$\Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
\label{thm:diameter}
\end{theorem}

\proof
Since $d_c([1,2,3,\ldots,n], [n-1,n,n-2,n-3, \ldots, 1]) = \frac{(n-1)(n-2)}{2}$, the claim follows directly from from Theorem~\ref{thm:caterpillar_convex} and Lemma~\ref{lemma:diameter_bound}.
\endproof

We show in the rest of this section that the radius of $\rnni$ space coincides with its diameter.
The main tool to establish this result is the algorithm $\mdtree$ from Section~\ref{section:alg_mdtree}.
Recall that $\mdtree$ is a non-deterministic algorithm, which implies that the output tree is not uniquely defined by the input to the algorithm.

\begin{lemma}
Let $T$ be a caterpillar tree.
Then $d(T, R) = \Delta(\rnni)$ for every tree $R$ such that $R = \mdtree(T)$.
\label{lemma:max_dist_caterpillar}
\end{lemma}

\begin{proof}
Recall that with Theorem~\ref{thm:diameter} we know that $\Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
Within this proof we assume, without loss of generality, that it is $T = [1,2,3,\ldots,n]$.
We prove the lemma by induction on the number of taxa $n$.

@@

In Section~\ref{alg:max_dist_tree} we already discussed that $\mdtree$ is a non-deterministic algorithm.
In fact, the output tree of $\mdtree$ can have any tree topology, since in each step of the algorithm the next taxon can be added on any edge incident to a leaf in the tree.
Therefore, if the input tree is a caterpillar tree, $\mdtree$ can compute a tree with any tree topology that has maximum distance from the input tree.
And as permuting labels of trees does not change the distance between them, there exists for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
By this we obtain Theorem~\ref{thm:radius}.

\begin{theorem}
The radius of the $\rnni$ graph is $\rad(\rnni) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
\label{thm:radius}
\end{theorem}
